The Unicode Cookbook \linebreak for Linguists

### 2015-12-15 book/chapters/orthography_profiles.tex

The major challenge then is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
a nasalized vowel /ṽ/).

@@

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       should then explicitly report on the order used.      
\end{enumerate}

@@

@@

[Here should a few abstract short simple examples be added]

\

>>>>

The major challenge then is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
a nasalized vowel /Ṽ/).

@@

The logic of orthographic parsing ...

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       should then explicitly report on the order used.      
\end{enumerate}

@@

\section{Available implementations}

@@

In the following example, it looks as if we have two identical strings,
\texttt{AABB}. However, this is just an surface-impression delivered by the
current font, which renders Latin and Cyrillic capitals identically. We can
identify this problem when we produce an orthography profile from the strings.
Using here the R implementation of orthography profiles, we first assign the two
strings to a variable \texttt{test}, and then produce an orthography profile
with the function \texttt{write.profile}. As it turns out, some of the letters 
are Cyrillic.

<<>>==
test <- c('AABB', '\u0041\u0410\u0042\u0412')
write.profile(test)
@

\noindent The working of error-message reporting can also nicely be illustrated with this
example. Suppose we made an orthography profile with just the Latin <A> and <B> 
as possible graphemes, then this profile would not be sufficient to tokenize the 
strings. There are graphemes in the data that are not in the profile, so the 
tokenization produces an error, which can be used to fix the encoding. In the 
example below, we can see that the Cyrillic encoding is found in the second 
string of the \texttt{test} input.

<<>>==
test <- c('AABB', '\u0041\u0410\u0042\u0412')
tokenize( test, profile = c('A', 'B') )
@
